Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-01-19 Multiresolution_framework/ltx/manuscript.tex
edits to Bspline section

The utility of these models for answering questions in clinical neuroscience and neurology lies in the ability the infer parameters from electrophysiological data. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are ideal candidates for estimation algorithms due to their strong links with physiology and their parsimony. It is expected that parameters of the neural fields models will be patient-specific and will, therefore, need to be inferred from data.

@@

The $m$-th  order cardinal B-spline function is defined by the recurrence relation \cite{Chui1992} 
\begin{equation}
N_{m}\left(r\right) = \left(N_{m-1}\ast N_{1}\right)\left(r\right) = \int_0^{1} N_{m-1}\left( r-r'\right)\,\mathrm{d}r',
\label{SplineConvolutionIntegral}
\end{equation}
where $m>1$, $\ast$ denotes convolution, and $N_1\left(r\right)$ is the characteristic function of the unit interval $\left[ 0,1\right)$, also known as indicator function, i.e.
\begin{equation}
N_{1}\left(r\right)=
\begin{cases}
1 & \text{if $ 0\le r<1$}, \\
0 & \mathrm{elsewhere}.
\end{cases}
\end{equation}
Following this, a B-spline of any order, $N_m(r)$, can be computed using the recursive expression defined by \cite{DeBoor2001}
\begin{equation}\label{eq:MRA-DoBoorFormula}
 N_{m}\left(r\right)=\frac{r}{m-1}N_{m-1}\left(r\right)+\frac{m-r}{m-1}N_{m-1}\left(r-1\right) \quad m>1.
 \end{equation}
The most commonly used form of splines is the cubic spline $\left(m=4\right)$, comprised of third order polynomials added together at the joining points, with an explicit expression given by
\begin{align}
N_{4}\left(r\right)=
\begin{cases}
r^3 & \text{if $ 0\le r\le1$}, \\
4-12r+12r^2-3r^3 & \text{if $1\le r\le2$}, \\
-44+60r-24r^2+3r^3 & \text{if $2\le r\le3$}, \\
64-48r+12r^2-r^3 & \text{if $3\le r\le4$}, \\
0 & \mathrm{elsewhere}.
\end{cases}
\end{align}
The convolution and inner product of B-spline functions are required in order to formulate the multi-resolution framework described in the previous section (construction of $\boldsymbol\Lambda_x$, $\boldsymbol\Lambda_{\theta}$ and $\boldsymbol\Sigma_w$). To show how the convolution of two B-splines is calculated, \eqref{SplineConvolutionIntegral} can be rewritten as $(m-1)$ convolutions of the indicator function with itself
\begin{equation}\label{eq:N1convolutions}
 N_{m}\left(r\right)=\underbrace{\left(N_{1}\ast N_{1}\ast \cdots \ast N_{1}\right)}_{m-1\quad \text{convolutions}}\left(r\right).
\end{equation}
Using the associativity property of convolution, we have
\setlength{\arraycolsep}{0.0em}
\begin{align}\label{eq:BsplineConvolution}
N_{m}\left( r\right) \ast N_{m'}\left(r\right)&=\underbrace{\overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m-1 \quad \text{convolutions}}\left(r\right) \ast \overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m'-1\quad \text{convolutions}}}_{m+m'-1 \quad \text{convolutions}}\left(r\right)\nonumber\\
&=N_{m+m'}\left(r\right).
\end{align}
A direct consequence of \eqref{eq:BsplineConvolution} is the inner product between two B-splines is another translated B-spline such that
\begin{align}
 \left\langle N_{m}\left(r-l_{1}\right), N_{m'}\left(r-l_{2}\right)\right\rangle=&N_{m+m'}\left(m+l_{1}-l_{2}\right)\nonumber \\
=&N_{m+m'}\left(m'+l_{2}-l_{1}\right),
\label{eq:BsplineInnerProduct}
\end{align}
where $\left\langle \cdot,\cdot\right\rangle $ denotes the inner product. This holds as the support of $N_m\left(r\right)$ is $\left[ 0,m\right]$ and  is symmetric with respect to $r=\frac{m}{2}$, i.e. $ N_{m}\left(\frac{m}{2}+r\right)=N_{m}\left(\frac{m}{2}-r\right)$ (see Appendix~\ref{ap:InnerProductOfBsplines} for explicit derivation). The values of $N_{m+m'}$ in \eqref{eq:BsplineInnerProduct} can be easily determined recursively by evaluating \eqref{eq:MRA-DoBoorFormula} at integer points, i.e.
 \begin{equation}\label{eq:MRA-recursiveBsplineatintegerpoints}
 \begin{cases}
 N_2(k)=\delta(k-1)\quad k\in \mathbb{Z}, \\
 N_{m}\left(k\right)=\frac{k}{m-1}N_{m-1}\left(k\right)+\frac{m-k}{m-1}N_{m-1}\left(k-1\right) \quad k=1,2,\dots,m-1
  \end{cases}
 \end{equation}
Note $N_{m}\left(k\right)=0$ for $k\le0$ or $k\ge m$.

The multi-resolution approximation using B-spline functions is completed by defining a two-scale relation pair, which for cardinal B-spline scaling and wavelet functions of order $m$ takes the form of
\begin{align}
 N_{m}\left(r\right)&=\sum_{n=0}^{m} \left[\mathbf p\right]_n N_{m}\left(2r-n\right) \label{eq:MRA-TwoScalepair1} \\
  \varphi_{m}\left(r\right) &= \sum_{n=0}^{3m-2} \frac{\left(-1\right)^n}{2^{m-1}} \left[\mathbf q\right]_n N_{m}\left(2r-n\right)\label{eq:MRA-TwoScalepair2},
 \end{align}
where 
 \begin{align}
\left[\mathbf p\right]_n&=2^{-m+1} \binom{m}{n} \quad \text{ $0\le n\le m$} \label{eq:MRA-TwoScalepair1coefs}\\
\left[\mathbf q\right]_n&= \sum_{l=0}^{m} \binom{m}{l} N_{2m}\left(n-l+1\right), \,  \text{ $0\le n\le 3m-2$}\label{eq:MRA-TwoScalepair2coefs}
 \end{align}
and $\left[\cdot\right]_n$ denotes the $n$-th element of the vector coefficients \cite{Chui1992}. \dean{Need to explicitly say why this is important and what this gives us here.}

@@

In order to analytically calculate elements of $\boldsymbol\Lambda_{x}$ and $\boldsymbol\Lambda_{\theta}$, the scaling and wavelet basis functions must be expanded in terms of $N_m$ at the appropriate scale. For scaling functions this can be done  by using \eqref{eq:MRA-TwoScalepair1} recursively, giving
\begin{align}\label{eq:MRA-ConvertingFormulaForscalingFunctions}
 &N_m(r)=\nonumber \\
&\sum_{n_1,n_2, \dots n_j=0}^{m}\left[\mathbf p\right]_{n_1} \left[\mathbf p\right]_{n_2}\dots \left[\mathbf p\right]_{n_j}N_m(2^jr-l_{n_1,n_2, \dots, n_j}),
\end{align}
where 
\begin{align}
 l_{n_1,n_2, \dots, n_j}=2^{j-1}n_1+2^{j-2}n_2+ \dots +2^{0}n_j.
\end{align}
The relation \eqref{eq:MRA-ConvertingFormulaForscalingFunctions} can be also proven by induction (see Appendix ...). A similar method can be used for wavelet functions noting that they must be written first in terms of scaling functions using \eqref{eq:MRA-TwoScalepair2}. Examples of such expansions are given in Figure~\ref{fig:MRA-BasisDecomposition} where the scaling and wavelet functions at $j=0$ are written in terms of 625 and 1250 scaling functions at $j=4$ respectively.
\begin{figure}[!t] 
 	\centering
 		\includegraphics[scale=1]{./Graph/Decomposition.pdf}
 		\caption{{\bf The scaling and the wavelet function decomposition}. (\textbf{A}) The scaling function (blue) at $j=0$ is obtained using 625 scaling functions (black) at $j=4$. (\textbf{B}) The wavelet function (blue) at $j=0$ is obtained using 1250 scaling functions (black) at $j=4$.}
 	\label{fig:MRA-BasisDecomposition}
 \end{figure}  
 In this paper, 4th -- order cardinal B-spline scaling and wavelet functions are used, resulting in  8th and 12th order B-spline functions when calculating \eqref{eq:Lambdax}, \eqref{eq:Lambdatheta} and \eqref{eq:CovMatrix} whose values at integer points are given in Table~\ref{table:MRA-BsplineatIntegerPoints}.
\begin {table}[t]
\begin{center}
\begin{tabular}{lcccc}
 & $k$& $7!N_{8}\left(k\right)$ & $11!N_{12}\left(k\right)$\\ \hline\hline
 & 1  & $1$ & $1$\\
 & 2  & $120$ & $2,036$\\
 & 3  & $1,191$ & $152,637$\\
 & 4  & $2,416$ & $2,203,488$\\
 & 5  & $\space$ & $9,738,114$\\
 & 6  & $\space$ & $15,724,248$
 \end{tabular}
 \caption {{\bf Cardinal B-Splines at the knot sequence}. Table entries are calculated using \eqref{eq:MRA-recursiveBsplineatintegerpoints} and can be also found in \cite{Goswami1999}.} 
 \label{table:MRA-BsplineatIntegerPoints}
 \end{center}
 \end {table}
\subsection{Frequency Analysis}
To determine the level of approximation in the model, detailed in the previous section, that is the number of basis functions required to reconstruct the field, $v_t(r)$, the frequency response of the B-spline function needs to be computed by applying the Fourier transform to \eqref{eq:N1convolutions}. To proceed the Fourier transform of $N_1(r)$ is first calculated, 
\begin{align}\label{eq:MRA-N1Fouriertransform}
\mathcal F(N_1(r))&=\int_{-\infty}^{+\infty}N_1(r)\mathrm{exp}(-2\pi i \nu r)dr \nonumber \\
&=\int_{0}^{1} \mathrm{exp}(-2\pi i \nu r)dr \nonumber \\
&=\frac{1-\mathrm{exp}(-2\pi i \nu)}{2\pi i\nu}.
\end{align}
The convolution theorem states that the Fourier transform of the convolution of functions is equal to the product of their Fourier transforms. Therefore, taking Fourier transform of \eqref{eq:N1convolutions} and substituting \eqref{eq:MRA-N1Fouriertransform} for $\mathcal F(N_1(r)) $ yields
\begin{align}\label{eq:MRA-NmFouriertransform}
\mathcal F(N_m(r))=\left(\frac{1-\mathrm{exp}(-2\pi i \nu)}{2\pi i\nu}\right)^m.
\end{align}
\dean{Need to say how the FT is used somewhere here before proceeding.}

@@

The stopping rule adopted herein is

>>>>

The utility of these models for answering questions in clinical neuroscience and neurology lies in the ability to infer parameters from electrophysiological data. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are ideal candidates for estimation algorithms due to their strong links with physiology and their parsimony. It is expected that parameters of the neural fields models will be patient-specific and will, therefore, need to be inferred from data.

@@

The $m$-th  order cardinal B-spline function is defined by the recurrence relation \cite{Chui1992} 
\begin{equation}
N_{m}\left(r\right) = \left(N_{m-1}\ast N_{1}\right)\left(r\right) = \int_0^{1} N_{m-1}\left( r-r'\right)\,\mathrm{d}r',
\label{SplineConvolutionIntegral}
\end{equation}
where $m>1$, $\ast$ denotes convolution, and $N_1\left(r\right)$ is the characteristic function of the unit interval $\left[ 0,1\right)$, also known as indicator function, i.e.
\begin{equation}
N_{1}\left(r\right)=
\begin{cases}
1 & \text{if $ 0\le r<1$}, \\
0 & \mathrm{elsewhere}.
\end{cases}
\end{equation}
Following this, a B-spline of any order, $N_m(r)$, can be computed using the recursive expression defined by \cite{DeBoor2001}
\begin{equation}\label{eq:MRA-DoBoorFormula}
 N_{m}\left(r\right)=\frac{r}{m-1}N_{m-1}\left(r\right)+\frac{m-r}{m-1}N_{m-1}\left(r-1\right) \quad m>1.
 \end{equation}
The most commonly used form of splines is the cubic spline $\left(m=4\right)$, comprised of third degree polynomials added together at the joining points, with an explicit expression given by
\begin{align}
N_{4}\left(r\right)=
\begin{cases}
r^3 & \text{if $ 0\le r\le1$}, \\
4-12r+12r^2-3r^3 & \text{if $1\le r\le2$}, \\
-44+60r-24r^2+3r^3 & \text{if $2\le r\le3$}, \\
64-48r+12r^2-r^3 & \text{if $3\le r\le4$}, \\
0 & \mathrm{elsewhere}.
\end{cases}
\end{align}
The convolution and inner product of B-spline functions are required in order to formulate the multi-resolution framework described in the previous section (construction of $\boldsymbol\Lambda_x$, $\boldsymbol\Lambda_{\theta}$ and $\boldsymbol\Sigma_w$). To show how the convolution of two B-splines is calculated, \eqref{SplineConvolutionIntegral} can be rewritten as $(m-1)$ convolutions of the indicator function with itself
\begin{equation}\label{eq:N1convolutions}
 N_{m}\left(r\right)=\underbrace{\left(N_{1}\ast N_{1}\ast \cdots \ast N_{1}\right)}_{m-1\quad \text{convolutions}}\left(r\right).
\end{equation}
Using the associativity property of convolution, we have
\setlength{\arraycolsep}{0.0em}
\begin{align}\label{eq:BsplineConvolution}
N_{m}\left( r\right) \ast N_{m'}\left(r\right)&=\underbrace{\overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m-1 \quad \text{convolutions}}\left(r\right) \ast \overbrace{\left(N_{1} \ast \cdots \ast N_{1}\right)}^{m'-1\quad \text{convolutions}}}_{m+m'-1 \quad \text{convolutions}}\left(r\right)\nonumber\\
&=N_{m+m'}\left(r\right).
\end{align}
A direct consequence of \eqref{eq:BsplineConvolution} is the inner product between two B-splines is another translated B-spline such that
\begin{align}
 \left\langle N_{m}\left(r-l_{1}\right), N_{m'}\left(r-l_{2}\right)\right\rangle=&N_{m+m'}\left(m+l_{1}-l_{2}\right)\nonumber \\
=&N_{m+m'}\left(m'+l_{2}-l_{1}\right),
\label{eq:BsplineInnerProduct}
\end{align}
where $\left\langle \cdot,\cdot\right\rangle $ denotes the inner product. This holds as the support of $N_m\left(r\right)$ is $\left[ 0,m\right]$ and  is symmetric with respect to $r=\frac{m}{2}$, i.e. $ N_{m}\left(\frac{m}{2}+r\right)=N_{m}\left(\frac{m}{2}-r\right)$ (see Appendix~\ref{ap:InnerProductOfBsplines} for explicit derivation). The values of $N_{m+m'}$ in \eqref{eq:BsplineInnerProduct} can be easily determined recursively by evaluating \eqref{eq:MRA-DoBoorFormula} at integer points, i.e.
 \begin{equation}\label{eq:MRA-recursiveBsplineatintegerpoints}
 \begin{cases}
 N_2(k)=\delta(k-1)\quad k\in \mathbb{Z}, \\
 N_{m}\left(k\right)=\frac{k}{m-1}N_{m-1}\left(k\right)+\frac{m-k}{m-1}N_{m-1}\left(k-1\right) \quad k=1,2,\dots,m-1
  \end{cases}
 \end{equation}
Note $N_{m}\left(k\right)=0$ for $k\le0$ or $k\ge m$.

The multi-resolution approximation using B-spline functions is completed by defining a two-scale relation pair, which relates the scaling functions and the wavelets at a given scale with the scaling function at the next higher scale. For cardinal B-spline scaling and wavelet functions of order $m$ the two-scale relation pair takes the form of
\begin{align}
 N_{m}\left(r\right)&=\sum_{n=0}^{m} \left[\mathbf p\right]_n N_{m}\left(2r-n\right) \label{eq:MRA-TwoScalepair1} \\
  \varphi_{m}\left(r\right) &= \sum_{n=0}^{3m-2} \frac{\left(-1\right)^n}{2^{m-1}} \left[\mathbf q\right]_n N_{m}\left(2r-n\right)\label{eq:MRA-TwoScalepair2},
 \end{align}
where 
 \begin{align}
\left[\mathbf p\right]_n&=2^{-m+1} \binom{m}{n} \quad \text{ $0\le n\le m$} \label{eq:MRA-TwoScalepair1coefs}\\
\left[\mathbf q\right]_n&= \sum_{l=0}^{m} \binom{m}{l} N_{2m}\left(n-l+1\right), \,  \text{ $0\le n\le 3m-2$}\label{eq:MRA-TwoScalepair2coefs}
 \end{align}
and $\left[\cdot\right]_n$ denotes the $n$-th element of the vector coefficients \cite{Chui1992}. \dean{Need to explicitly say why this is important and what this gives us here.}

@@

In order to analytically calculate elements of $\boldsymbol\Lambda_{x}$ and $\boldsymbol\Lambda_{\theta}$, the scaling and wavelet basis functions must be expanded in terms of $N_m$ at the appropriate scale using the two-scale relation pair defined in equations (\ref{eq:MRA-TwoScalepair1}-\ref{eq:MRA-TwoScalepair2coefs}). For scaling functions this can be done  by using \eqref{eq:MRA-TwoScalepair1} recursively, giving
\begin{align}\label{eq:MRA-ConvertingFormulaForscalingFunctions}
 &N_m(r)=\nonumber \\
&\sum_{n_1,n_2, \dots n_j=0}^{m}\left[\mathbf p\right]_{n_1} \left[\mathbf p\right]_{n_2}\dots \left[\mathbf p\right]_{n_j}N_m(2^jr-l_{n_1,n_2, \dots, n_j}),
 l_{n_1,n_2, \dots, n_j}=2^{j-1}n_1+2^{j-2}n_2+ \dots +2^{0}n_j.
The relation \eqref{eq:MRA-ConvertingFormulaForscalingFunctions} can be also proven by induction (see Appendix ...). A similar method can be used for wavelet functions noting that they must be written first in terms of scaling functions using \eqref{eq:MRA-TwoScalepair2}. Examples of such expansions are given in Figure~\ref{fig:MRA-BasisDecomposition} where the scaling and wavelet functions at $j=0$ are written in terms of 625 and 1250 scaling functions at $j=4$ respectively.
 		\includegraphics[scale=1]{./Graph/Decomposition.pdf}
 		\caption{{\bf The scaling and the wavelet function decomposition}. (\textbf{A}) The scaling function (blue) at $j=0$ is obtained using 625 scaling functions (black) at $j=4$. (\textbf{B}) The wavelet function (blue) at $j=0$ is obtained using 1250 scaling functions (black) at $j=4$.}
 	\label{fig:MRA-BasisDecomposition}
 In this paper, 4th -- order cardinal B-spline scaling and wavelet functions are used, resulting in  8th and 12th order B-spline functions when calculating \eqref{eq:Lambdax}, \eqref{eq:Lambdatheta} and \eqref{eq:CovMatrix} whose values at integer points are given in Table~\ref{table:MRA-BsplineatIntegerPoints}.
\begin {table}[t]
\begin{tabular}{lcccc}
 & $k$& $7!N_{8}\left(k\right)$ & $11!N_{12}\left(k\right)$\\ \hline\hline
 & 1  & $1$ & $1$\\
 & 2  & $120$ & $2,036$\\
 & 3  & $1,191$ & $152,637$\\
 & 4  & $2,416$ & $2,203,488$\\
 & 5  & $\space$ & $9,738,114$\\
 & 6  & $\space$ & $15,724,248$
 \end{tabular}
 \caption {{\bf Cardinal B-Splines at the knot sequence}. Table entries are calculated using \eqref{eq:MRA-recursiveBsplineatintegerpoints} and can be also found in \cite{Goswami1999}.} 
 \label{table:MRA-BsplineatIntegerPoints}
\subsection{Frequency Analysis}
To determine the level of approximation in the model, detailed in the previous section, that is the number of basis functions required to reconstruct the field, $v_t(r)$, the frequency response of the B-spline function needs to be computed by applying the Fourier transform to \eqref{eq:N1convolutions}. To proceed the Fourier transform of $N_1(r)$ is first calculated, 
\begin{align}\label{eq:MRA-N1Fouriertransform}
\mathcal F(N_1(r))&=\int_{-\infty}^{+\infty}N_1(r)\mathrm{exp}(-2\pi i \nu r)dr \nonumber \\
&=\int_{0}^{1} \mathrm{exp}(-2\pi i \nu r)dr \nonumber \\
&=\frac{1-\mathrm{exp}(-2\pi i \nu)}{2\pi i\nu}.
The convolution theorem states that the Fourier transform of the convolution of functions is equal to the product of their Fourier transforms. Therefore, taking Fourier transform of \eqref{eq:N1convolutions} and substituting \eqref{eq:MRA-N1Fouriertransform} for $\mathcal F(N_1(r)) $ yields
\begin{align}\label{eq:MRA-NmFouriertransform}
\mathcal F(N_m(r))=\left(\frac{1-\mathrm{exp}(-2\pi i \nu)}{2\pi i\nu}\right)^m.
\dean{Need to say how the FT is used somewhere here before proceeding.} \parham{The Fourier transform of the $m$-th order Bspline allows the calculation of the spatial frequency response of the wavelets using \eqref{eq:MRA-TwoScalepair2} from which the level of approximation, $j$, can be determined. This way we choose wavelets upto level $j$ which fully describe the spatial frequency contents of the underlying neural field estimated from observations.}

@@

The stopping criterion is usually associated with either the change in the parameter estimates or the log-likelihood variation \cite{McLachlan1997}. The stopping rule adopted herein is
